Characteristic polynomials of p-adic matrices

### 2017-04-21 charpoly/charpoly.tex

We recall some of the definitions and results of \cite{caruso-roe-vaccon:14a}
as a foundation for our discussion of the precision for the characteristic polynomial
of a matrix.  We will be concerned with two $K$-manifolds in what follows:
the space $M_n(K)$ of $n \times n$ matrices with entries in $K$ and the space
$K_n[X]$ of monic degree $n$ polynomials over $K$.  One cannot represent
matrices $M \in M_n(K)$ exactly, but instead must specify an approximation
together with a measure of the approximation's accuracy.  Recall that a
\emph{lattice} in $M_n(K)$ is an open $\OK$-submodule of $M_n(K)$:
the $\OK$-span of $n^2$ basis vectors \cite{caruso-roe-vaccon:14a}*{\S 2.2}.
We record the precision of $M$ with lattice $H$: the matrix $M$ is understood
as some matrix of the form $M + h$, where $h \in H$.

One problem with representing precision in this way is that storing $H$
requires $n^2$ basis vectors, each with $n^2$ entries.  We can shrink the
space required by working only with certain classes of lattices.  When
using \emph{jagged} precision, the basis matrices for $H$ can only have
one nonzero entry; this corresponds to specifying a precision for each
entry of the matrix separately.  Going further, $M$ is said to have \emph{flat}
precision $O(p^N)$ if every entry has precision $O(p^N)$, or equivalently,
if $H = p^N M_n(\OK)$.  Similarly, precision for
monic polynomials can be specified by giving a lattice in $K_{<n}[X]$,
or specifying precision coefficient-wise in the interest of simplicity.

Let $\chi : M_n(K) \to K_n[X]$ be the characteristic polynomial map.
Our analysis of the precision behavior of $\chi$ rests upon
the computation of its derivative $d\chi$, using \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.
Let $\adj(M)$ denote the adjugate\footnote{In \citelist{\cite{caruso-roe-vaccon:14a} \cite{caruso-roe-vaccon:15a}} the adjugate was referred to using the non-standard term \emph{comatrix}} of $M$ (when $M \in \GL_n(K)$,
we have $\adj(M) = \det(M) M^{-1}$) and $d\chi_M$ the differential at $M$.  Recall
\citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} \cite{caruso-roe-vaccon:15a}*{\S 3.3}}
that $d\chi_M$ is given by
\begin{equation} \label{eq:dchi}
d\chi_M: dM \mapsto \tr(\adj(X-M) \cdot dM).
\end{equation}

When $d\chi_M$ is surjective and $H$ is small enough, we may determine the precision of $\chi(M)$ precisely.
Write $V$ for $M_n(K)$, $W$ for $K_n[X]$, 
$B^-_V(r)$ (resp. $B_V(r)$) for the open (resp. closed) ball of radius $r$ in $V$, and let
$\sigma_1(M), \dots, \sigma_n(M)$ denote the elementary divisors of $M$.

\begin{prop} \label{prop:mainlem}
Let $M \in M_n(K)$ and suppose that $d\chi_M$ is surjective.  Let 
\[
\alpha = \min\left(\prod_{i=1}^{n-1} \left\vert \sigma_i(M) \right\vert, 1\right).
\]
Then, for all $\rho \in (0, 1]$ and all
$r \in (0, \alpha^{-1} \cdot \rho^{-1})$, any lattice $H$ such that $B_V^-(\rho r) \subset H \subset B_V(r)$
satisfies:
\begin{equation}
\chi(M + H) = \chi(M) + d\chi_M(H).
\end{equation}
\end{prop}

\begin{proof}
Recall \cite{caruso-roe-vaccon:15a}*{Def. 3.3} that the \emph{precision polygon}
of $M$ is the lower convex hull of the Newton polygons of the entries of $\adj(X-M)$.
By \cite{caruso-roe-vaccon:15a}*{Prop. 3.4}, the endpoints of the precision polygon
occur at height $0$ and $\sum_{i=1}^{n-1} \val(\sigma_i(M))$.  By convexity,
$B_W(1) \subset d\chi_M(B_V(\alpha^{-1}))$.

@@

We may rephrase the condition that $d\chi_M$ is surjective using more familiar invariants attached to $M$.
\begin{prop}
\label{prop:surjectivity}
For $M \in M_n(K)$, the following conditions are equivalent:
\begin{enumerate}[(i)]
\renewcommand{\itemsep}{0pt}
\item \label{pt:surj} the differential $d\chi_M$ is surjective
\item \label{pt:cyc} the matrix $M$ has a cyclic vector (\emph{i.e.} $M$ is similar
to a companion matrix)
\item \label{pt:1d} the eigenspaces of $M$ over the algebraic
closure $\bar{K}$ of $K$ all have dimension $1$
\item \label{pt:charmin} the characteristic polynomial of $M$ is equal to the minimal polynomial of $M$.
\end{enumerate}
\end{prop}

@@

\begin{prop}
\label{prop:intsurj}
For $M \in M_n(\OK)$, the following conditions are equivalent:
\begin{enumerate}[(i)]
\renewcommand{\itemsep}{0pt}
\item \label{pt:surjp} the image of $M_n(\OK)$ under $d\chi_M$ is $\OK[X] \cap K_{<n}[X]$.
\item \label{pt:cycp} the reduction of $M$ modulo $\pi$ has a cyclic vector.
\end{enumerate}
\end{prop}
\begin{proof}
The condition \eqref{pt:surjp} is equivalent to the surjectivity of $d\chi_M$ modulo $\pi$.  The equivalence with
\eqref{pt:cycp} follows the same argument as Proposition \ref{prop:surjectivity}, but over the residue field of $K$.
\end{proof}

@@

\begin{rem}
With the previous notation, if $U \in GL_n(K),$
the quadruple for 
$UMU^{-1}$ is
$(\alpha, UP, (U^T)^{-1}Q, \chi_M),$
which can be computed in $O(n^\omega)$ operations in $K.$
This is faster than computing $U \adj(X-M) U^{-1},$
which is, at first sight, in $O(n^4)$ operations in $K.$
\end{rem}

>>>>

We recall some of the definitions and results of \cite{caruso-roe-vaccon:14a}
as a foundation for our discussion of the precision for the characteristic polynomial
of a matrix.  We will be concerned with two spaces in what follows:
the space $M_n(K)$ of $n \times n$ matrices with entries in $K$ and the space
$K_n[X]$ of monic degree $n$ polynomials over $K$.  We also define $K_{<n}[X]$ (resp. 
${\OK}_{<n}[X]$) as the space of polynomials over $K$ (resp. over $\OK$) 
of degree strictly less than $n$.

One cannot represent
matrices $M \in M_n(K)$ exactly, but instead must specify an approximation
together with a measure of the approximation's accuracy; there are several ways to do so.
When using \emph{jagged} precision, on gives a precision on each entry,
representing the $(i,j)$ entry as $a_{i,j} + O(\pi^{N_{i,j}})$. The integer $N_{i,j}$ 
is called the \emph{absolute precision} and may depend on $(i,j)$.
When the $N_{i,j}$ are all equal, we say that the precision datum is 
\emph{flat}.

We may also model precision more flexibly by using lattices.  We 
recall that a \emph{lattice} $H$ in $M_n(K)$ is the $\OK$-span of a vector
space basis, or equivalently the image of $M_n(\OK)$ under a 
$K$-linear transformation of $M_n(K)$ \cite{caruso-roe-vaccon:14a}*{\S 2.2}.
For $M \in M_n(K)$, cosets of the form $M + H$ are special
neighborhoods of $M$; we will use them for modeling precision on $M$. 
Given a family of integers $\famN = (N_{i,j})_{i,j}$ the lattice
\[
H_{\famN} = \big\{ \, M = (a_{i,j})_{i,j} \in M_n(\Qp) \,\, | \,\, \forall i,\!j\!:
\pi^{N_{i,j}} \text{ divides } a_{i,j} \, \big\}
\]
corresponds to the jagged precision introduced above.
The benefit of working with lattices is that they behave well under 
differentiable maps \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}, but they
require more space and time to compute.

Jagged and flat precision is defined similarly for monic polynomials, 
taking care that no precision is specified for the leading 
coefficient which is supposed to be exactly~$1$. The notion 
of \emph{lattice precision} also extends to polynomials. The underlying 
vector space is now $K_{<n}[X]$ and we define a lattice in it as the
image of ${\OK}_{<n}[X]$ under some linear mapping.

Let $\chi : M_n(K) \to K_n[X]$ be the characteristic polynomial map
and $d \chi_M$ for its differential at $M$.
Our analysis of the precision behavior of $\chi$ rests upon
the computation of its derivative $d\chi$, using \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.
Let $\adj(M)$ denote the adjugate\footnote{In \citelist{\cite{caruso-roe-vaccon:14a} \cite{caruso-roe-vaccon:15a}} the adjugate was referred to using the non-standard term \emph{comatrix}} of $M$ (when $M \in \GL_n(K)$,
we have $\adj(M) = \det(M) M^{-1}$).  Recall
\citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} \cite{caruso-roe-vaccon:15a}*{\S 3.3}}
that $d\chi_M$ is given by
\begin{equation} \label{eq:dchi}
\begin{array}{rcl}
d\chi_M: \quad M_n(K) & \to & K_{<n}[X] \smallskip \\
 dM & \mapsto & \tr(\adj(X{-}M) \cdot dM).
\end{array}
\end{equation}

When $d\chi_M$ is surjective and $H$ is sufficiently ``small'' and ``regular,'' we may determine
the precision of $\chi(M)$ precisely as
\begin{equation} \label{eq:mainlem}
\chi(M + H) = \chi(M) + d\chi_M(H).
\end{equation}
More precisely, let
$\sigma_1(M), \dots, \sigma_n(M)$ denote the elementary divisors of $M$.

\begin{prop} \label{prop:mainlem}
Let $M \in M_n(K)$ and assume that $d \chi_M$ is surjective. Let:
\[
\nu = \max\left(\sum_{i=1}^{n-1} \val(\sigma_i(M)), 0\right).
\]
Then, for all integers $a, b \geq 0$ with $a + \nu < 2b$,
any lattice $H$ such that $M_n(\pi^b \OK) \subset H \subset
M_n(\pi^a \OK)$
satisfies Eq.~\eqref{eq:mainlem}
\end{prop}

\begin{prop} \label{prop:mainlem}
Let $M \in M_n(K)$ and assume that $d\chi_M$ is surjective.  Let 
\[
\alpha = \min\left(\prod_{i=1}^{n-1} \left\vert \sigma_i(M) \right\vert, 1\right).
\]
Then, for all $\rho \in (0, 1]$ and all
$r \in (0, \alpha^{-1} \cdot \rho^{-1})$, any lattice $H$ such that $B_V^-(\rho r) \subset H \subset B_V(r)$
satisfies:
\begin{equation}
\chi(M + H) = \chi(M) + d\chi_M(H).
\end{equation}
\end{prop}

\begin{proof}
Recall \cite{caruso-roe-vaccon:15a}*{Def. 3.3} that the \emph{precision polygon}
of $M$ is the lower convex hull of the Newton polygons of the entries of $\adj(X-M)$.
By \cite{caruso-roe-vaccon:15a}*{Prop. 3.4}, the endpoints of the precision polygon
occur at height $0$ and $\sum_{i=1}^{n-1} \val(\sigma_i(M))$.  By convexity,
${\OK}_{<n}[X] \subset d\chi_M(M_n(\pi^\nu \OK))$.

@@

We may rephrase the condition that $d\chi_M$ is surjective using more familiar invariants attached to $M$.

\begin{prop}
\label{prop:surjectivity}
For $M \in M_n(K)$, the following conditions are equivalent:
\begin{enumerate}[(i)]
\renewcommand{\itemsep}{0pt}
\item \label{pt:surj} the differential $d\chi_M$ is surjective
\item \label{pt:cyc} the matrix $M$ has a cyclic vector (\emph{i.e.} $M$ is similar
to a companion matrix)
\item \label{pt:1d} the eigenspaces of $M$ over the algebraic
closure $\bar{K}$ of $K$ all have dimension $1$
\item \label{pt:charmin} the characteristic polynomial of $M$ is equal to the minimal polynomial of $M$.
\end{enumerate}
\end{prop}

@@

\begin{prop}
\label{prop:intsurj}
For $M \in M_n(\OK)$, the following conditions are equivalent:
\begin{enumerate}[(i)]
\renewcommand{\itemsep}{0pt}
\item \label{pt:surjp} the image of $M_n(\OK)$ under $d\chi_M$ is 
${\OK}_{<n}[X]$.
\item \label{pt:cycp} the reduction of $M$ modulo $\pi$ has a cyclic vector.
\end{enumerate}
\end{prop}
\begin{proof}
The condition \eqref{pt:surjp} is equivalent to the surjectivity of $d\chi_M$ modulo $\pi$.  The equivalence with
\eqref{pt:cycp} follows the same argument as Proposition \ref{prop:surjectivity}, but over the residue field of $K$.
\end{proof}

@@

\begin{rem}
With the previous notation, if $U \in GL_n(K),$
the quadruple for 
$UMU^{-1}$ is
$(\alpha, UP, (U^T)^{-1}Q, \chi_M),$
which can be computed in $O(n^\omega)$ operations in $K.$
This is faster than computing $U \adj(X-M) U^{-1},$
which is, at first sight, in $\softO(n^4)$
%\tristan{Shouldn't it be $\softO (n^4)$?}
operations in $K.$
\end{rem}
